DynaProg for Scala

### 2013-01-17 docs/report/benchmarks.tex
More report.

\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
We first define a signature that can fit both algebrae, then we specify for each algebra the functions. In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).

We also want to emphasize how bad this example is, with respect to the result construction, as every intermediate result would contain its whole backtrack, hereby duplicating the information multiple times. To overcome this issue {\color{red} XXX}
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignSignature extends Signature {
  type Alphabet = Char
  type Answer = (Int,String,String) // score, seq1, seq2
  override val h = maxBy((x:Answer)=>x._1)

  def gap(current:Int, g:(Int,Int)):Int
  def pair(a:Alphabet,b:Alphabet):Int
}

trait SmithWatermanAlgebra extends SeqAlignSignature {
  private val open = -3
  private val extend = -1
  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, score + ( open + (size-1) * extend )) }
  def pair(a:Char,b:Char) = if (a==b) 10 else -3
}

trait NeedlemanWunschAlgebra extends SeqAlignSignature {
  private val open = -15
  private val extend = -1
  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; score + ( open + (size-1) * extend ) }
  def pair(a:Char,b:Char) = if (a==b) 4 else -3
}
\end{lstlisting}

\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
  private def prettyGap(sw:Subword,in:Function1[Int,Char]):(String,String) = {
    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
  }

  val alignment:Tabulate = tabulate("M",(
    empty                         ^^ { _ => (0,".",".") }
  | seq1() -~ alignment           ^^ { case (g,(score,s1,s2)) => val (g1,g2)=prettyGap(g,in1); (gap(score,g),s1+g1,s2+g2) }
  |           alignment ~- seq2() ^^ { case ((score,s1,s2),g) => val (g1,g2)=prettyGap(g,in2); (gap(score,g),s1+g2,s2+g1) }
  | el1    -~ alignment ~- el2    ^^ { case (c1,((score,s1,s2),c2)) => (score+pair(c1,c2),s1+c1, s2+c2) }
  ) aggregate h)

  val axiom = alignment

  def align(s1:String,s2:String) = parse(s1.toArray,s2.toArray).head
}

object SeqAlign extends App {
  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra   // Smith-Waterman
  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra // Needleman-Wunsch
  val seq1 = "CGATTACA"
  val seq2 = "CCCATTAGAG"

  val (swScore,sw1,sw2) = SWat.align(seq1,seq2)
  println("Smith-Waterman alignment\n- Score: "+swScore+"\n- Seq1: "+sw1+"\n- Seq2: "+sw2+"\n")
  val (nwScore,nw1,nw2) = NWun.align(seq1,seq2)
  println("Needleman-Wunsch alignment\n- Score: "+nwScore+"\n- Seq1: "+nw1+"\n- Seq2: "+nw2+"\n")
}
\end{lstlisting}

>>>>

\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
First define a signature that can fit both algebrae, then specify for each algebra the functions. In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}). A naive idea would be to construct the alignment string immediately in the forward phase (that is immediately in the {\tt Answer}). However, is not efficient at all, both in terms of running time and space complexity perspective because substring are duplicated for every cell. The correct way to solve this issue is to use backtracking and forward construct these strings, as described below:
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignSignature extends Signature {
  type Alphabet = Char
  def start(x:Unit):Answer
  def gap1(g:(Int,Int),a:Answer):Answer
  def gap2(a:Answer,g:(Int,Int)):Answer
  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
}

trait SmithWatermanAlgebra extends SeqAlignSignature {
  type Answer = Int
  override val h = max[Int] _
  private val open = -3
  private val extend = -1
  def start(x:Unit) = 0
  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, a + ( open + (size-1) * extend )) }
  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 10 else -3)
}

trait NeedlemanWunschAlgebra extends SeqAlignSignature {
  type Answer = Int
  override val h = max[Int] _
  private val open = -15
  private val extend = -1
  def start(x:Unit) = 0
  def gap1(g:(Int,Int),a:Answer) = gap2(a,g)
  def gap2(a:Answer,g:(Int,Int)) = { val size=g._2-g._1; a + ( open + (size-1) * extend ) }
  def pair(c1:Alphabet,a:Answer,c2:Alphabet) = a + (if (c1==c2) 4 else -3)
}
\end{lstlisting}

\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqPrettyPrint extends SeqAlignSignature {
  type Answer = (String,String)

  def in1(k:Int):Alphabet
  def in2(k:Int):Alphabet
  private def prettyGap(sw:(Int,Int),in:Function1[Int,Char]):(String,String) = {
    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
  }

  def start(x:Unit) = (".",".")
  def gap1(g:(Int,Int),a:Answer) = { val (g1,g2)=prettyGap(g,in1); (a._1+g1,a._2+g2) }
  def gap2(a:Answer,g:(Int,Int)) = { val (g2,g1)=prettyGap(g,in2); (a._1+g1,a._2+g2) }
  def pair(c1:Char,a:Answer,c2:Char) = (a._1+c1,a._2+c2)
}

trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
  val alignment:Tabulate = tabulate("M",(
    empty                         ^^ start
  | seq1() -~ alignment           ^^ gap1
  |           alignment ~- seq2() ^^ gap2
  | el1    -~ alignment ~- el2    ^^ pair
  ) aggregate h)

  val axiom = alignment
}

object SeqAlign extends App {
  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra
  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra
  object pretty extends SeqAlignGrammar with SeqPrettyPrint
  val seq1 = "CGATTACA"
  val seq2 = "CCCATTAGAG"

  def align(name:String,s1:String,s2:String,g:SeqAlignGrammar) = {
    val (score,bt) = g.backtrack(s1.toArray,s2.toArray).head
    val (a1,a2) = pretty.build(s1.toArray,s2.toArray,bt)
    println(name+" alignment\n- Score: "+score+"\n- Seq1: "+a1+"\n- Seq2: "+a2+"\n")
  }

  align("Smith-Waterman",seq1,seq2,SWat)
  align("Needleman-Wunsch",seq1,seq2,SWat)
}
\end{lstlisting}
